---
title: "Practical Machine Learning Project - Predict Body Movement Analysis Report"
author: "YeongWei"
date: "Tuesday, January 19, 2016"
output: 
  html_document:
        fig_height: 9
        fig_width: 9
---

# Introduction

Devices such as *Jawbone UP*, *Nike FuelBand* and *Fitbit* are capable of collecting mass amount of body movement data relatively inexpensively. The data used  here are collected from 6 volunteered participants for performing barbell lifts in 6 different ways (each denoted by alphabet and treated as classification). The main objective of this project is to build prediction models based on the data mentioned, then use them to predict the classification (outcome) based on a given testing data set. 

The classifications / outcomes are as below,

1. Class A - exactly according to the specification
2. Class B - throwing the elbows to the front
3. Class C - lifting the dumbbell only halfway
4. Class D - lowering the dumbbell only halfway
5. Class E - throwing the hips to the front

More information about the experiment can be found at [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har).

# User Defined Function

The section consist the various user defined function definitions and descriptions that are used in the [modeling](#modeling) section. The purpose is to make the source code in [modeling](#modeling) section more concise.

### downloadFile

If file is not found in destination path then proceed to download.
```{r}
downloadFile <- function(url, file, destination="./") {
  if (!file.exists(paste(destination, file, sep=""))) {
      download.file(url=paste(url, file, sep=""), 
                    dest=paste(destination, file, sep=""), method="curl")
  }
}
```
### loadOrInstallPackage

If package is not installed then proceed to installation then followed by loading into environment.
```{r}
loadOrInstallPackage <- function(packageName, ret=TRUE) {
  findPackage <- function(packageName)
    packageName %in% rownames(installed.packages())
  
  if (!findPackage(packageName)) install.packages(packageName)
  if (findPackage(packageName)) 
    suppressWarnings(suppressMessages(library(packageName, character.only=TRUE)))
  if (ret) found
}
```

### purgeNAColumns

If column has 80% or more of *NA* values then drop column.
```{r}
purgeNAColumns <- function(dataFrame, purgableThreshold=0.8) {
  dataFrame[, ! colSums(is.na(dataFrame)) / nrow(dataFrame) >= purgableThreshold]
}
```

### obviousColumnReduction

Drop columns based on column names.
```{r}
obviousColumnReduction <- function(dataFrame, columns) {
  dataFrame[, ! (names(dataFrame) %in% columns)]
}
```

### rationalizeFactorColumns

If a column in *dataFrame1* is of type *factor* then make the same column in *dataFrame2* as *factor*.
```{r}
rationalizeFactorColumns <- function(dataFrame1, dataFrame2) {
  for (colName in names(dataFrame1)) {
    if (is.factor(dataFrame1[, colName])) {
      dataFrame2[, colName] <- as.factor(dataFrame2[, colName])
    }
  }
  dataFrame2
}
```

### readCsv

A wrapper for *read.csv*.
```{r}
readCsv <- function(pathToFile, headerFlag=TRUE, naStrings=c("NA")) {
  read.csv(file=pathToFile, header=headerFlag, na.strings=naStrings)
}
```

# Library

This section attempts to load the required packages with [`loadOrInstallPackage`](#loadorinstallpackage).
```{r}
loadOrInstallPackage("caret", FALSE)
loadOrInstallPackage("rpart", FALSE)
loadOrInstallPackage("randomForest", FALSE)
loadOrInstallPackage("parallel", FALSE)
loadOrInstallPackage("doParallel", FALSE)
loadOrInstallPackage("corrplot", FALSE)
loadOrInstallPackage("rpart.plot", FALSE)
```
# Seed

This section attempts to set the `seed` value.
```{r}
set.seed(212121)
```

# Data Processing

This section attempts to retrieve and prepare the data.
```{r}
downloadUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/"
trainingDataFile <- "pml-training.csv"
testingDataFile <- "pml-testing.csv"
```

## Downloading Data

This section attempts to download CSV files with [`downloadFile`](#downloadfile).
```{r}
downloadFile(downloadUrl, trainingDataFile)
downloadFile(downloadUrl, testingDataFile)
```

## Cleaning Data

This section attempts clean the downloaded data.

1. Assuming some values to be *NA* with the [`readCsv`](#readcsv) evaluation.
```{r}
naStrings <- c("NA", "#DIV/0!", " ", "")
trainingDataFrame <- readCsv(trainingDataFile, TRUE, naStrings)
testingDataFrame <- readCsv(testingDataFile, TRUE, naStrings)
```

2. Dropping columns that has 80% of *NA* values with [`purgeNAColumns`](#purgenacolumns).
```{r}
trainingDataFrame <- purgeNAColumns(trainingDataFrame)
testingDataFrame <- purgeNAColumns(testingDataFrame)
```

3. Droping unnecessary columns with [`obviousColumnReduction`](#obviouscolumnreduction).
```{r}
obviousReducibleColumns <- c("X", "user_name", "raw_timestamp_part_1",
                             "raw_timestamp_part_2", "cvtd_timestamp",
                             "new_window", "num_window")

trainingDataFrame <- obviousColumnReduction(trainingDataFrame, 
                                             obviousReducibleColumns)
testingDataFrame <- obviousColumnReduction(testingDataFrame, 
                                           obviousReducibleColumns)

```
# Modeling

This sections attempts to build 2 prediction models based on *Decision Tree* and *Random Forest* algorithm.

## Create Training and Testing Dataset

This section attemtps to construct the training(60%) and testing(40%) dataset based on data from the [cleaned data](#cleaningdata).
```{r}
inTrain <- createDataPartition(y=trainingDataFrame$classe, p=0.6, list=FALSE)
training <- trainingDataFrame[inTrain, ]
testing <- trainingDataFrame[-inTrain, ]
```

## Create Control Parameter for Model Training

The control parameter below attempts to exploit the parallel processing feature by creating processing cluster based on the number of CPU cores available.
```{r}
cluster <- makeCluster(detectCores() -1)
registerDoParallel(cluster)
fitControl <- trainControl(method="cv", number=5, allowParallel=TRUE)
```
## Prediction Model with Decision Tree

Training the model with the *rpart* method.
```{r cache=TRUE}
modelFitDecisionTree <- train(classe ~ ., method="rpart", data=training, trControl=fitControl)
predictionDecisionTree <- predict(modelFitDecisionTree, testing)
cmDecisionTree <- confusionMatrix(testing$classe, predictionDecisionTree)
cmDecisionTree
```

## Prediction Model with Random Forest

Training the model with *rf* method.
```{r cache=TRUE}
modelFitRandomForest <- train(classe ~ ., method="rf", data=training, trControl=fitControl)
predictionRandomForest <- predict(modelFitRandomForest, testing)
cmRandomForest <- confusionMatrix(testing$classe, predictionRandomForest)
cmRandomForest
```

```{r echo=FALSE}
stopCluster(cluster)
```

## Analysis between Prediction Models

This section evaluates the overall accuracy and estimate the out-of-sample error. 

Helper function below computes the estimated out-of-sample error,
```{r}
computeFunc <- function (value) {
  1 - as.numeric(value)
}
```

Helper function returns the argument,
```{r}
retFunc <- function(arg) {
  arg
}
```

Helper function below returns the percentagized string,
```{r}
percentagize <- function(value, func) {
  paste(round(func(value) * 100, 2), "%")
}
```

### Decision Tree
1. The overall accuracy is,
```{r}
percentagize(cmDecisionTree$overall[1], retFunc)
```

2. The estimated out-of-sample error is,
```{r}
percentagize(cmDecisionTree$overall[1], computeFunc)
```

### Random Forest
1. The overall accuracy is,
```{r}
percentagize(cmRandomForest$overall[1], retFunc)
```

2. The estimated out-of-sample error is,
```{r}
percentagize(cmRandomForest$overall[1], computeFunc)
```
# Conclusion / Result

Based on the [Modeling](#modeling) section, the prediction model with [Random Forest](#random-forest) is better than the [Decision Tree](#decision-tree) algorithm.

Therefore the [Random Forest](#randomforest) prediction model is used to predict the outcome (classe column) for the `testingDataFrame` test data from the [Cleaning Data](#cleaning-data) section.

```{r}
prediction <- predict(modelFitRandomForest, testingDataFrame)
```

The predicted outcomes are arranged as follows,
```{r}
predictionDf <- data.frame(index=c(1: length(prediction)), prediction=prediction)
predictionDf
```

# Appendix

1. Data exploration effort for `trainingDataFrame` from Cleaning Data](#cleaning-data) section for correlation hints.
```{r fig.width=10}
corrPlot <- cor(trainingDataFrame[, -length(names(trainingDataFrame))])
corrplot(corrPlot, method="color")
```

2. Decision Tree visualization for [Decision Tree](#decision-tree) prediction model.
```{r}
prp(modelFitDecisionTree$finalModel)
```